Cheminformatics, the Computer Science of Chemical Discovery, Turning Open-Source

### 2012-02-06 resubmit/cheminformatics_CACM_revision.tex
added resubmission to new dir;

\title{Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

@@

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: the
  interdisciplinary collaboration between computer scientists and
  molecular biologists that led as one of its major accomplishments to the
  sequencing of the human genome. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce
  new therapeutic molecules.  Only within the last decade -- as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, as a result of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Novel technologies in the life sciences churn out information at an
ever increasing rate -- public data stores such as that of the
European Bioinformatics Institute (EBI) contain in the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available software implementing many of
these algorithms has only been accessible in the last ten to fifteen
years \cite{faulon2010}.

@@

Given that the art and science of drug discovery is based around small
molecules, techniques and algorithms to handle, analyse and interpret
chemical structure information are a core component of the drug
discovery toolbox. Public availability of chemical information and
tools to handle it are important as drug discovery research is slow,
expensive and prone to failure. The more information available to each
researcher, facilitating the development of novel treatments, the
better the chances of avoiding the many causes of attrition. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
\emph{annotations} (comments on the structure that are database
searchable), such as synonyms (different names for the same molecule),
known targets (drugs with which the molecule is known to interact),
mode of action (how the molecule interacts with its targets),
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

@@

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The representations range
from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
much implicit information (such as unlabeled nodes represent carbons
with valencies satisified by hydrogens) to a fully labeled graph,
where all information is explicit
(Fig.~\ref{figure:chemical-structures}B). The former is what is
usually exchanged informally between chemists, but to actually mine
chemical structures computationally, something like the latter is
required.  For example, graph mining algorithms can be applied to
labeled graph representations, to discover patterns in large-scale
datasets \cite{horst2009}.

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (A), a labeled graph (B), a 3D
  conformation (C) and a surface represention (D). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes, with the choice of representation guided by its
  intended use.}
\label{figure:chemical-structures}
\end{figure}

@@

Our motivation for this article is to highlight a variety of topics in
cheminformatics, presenting the computational and algorithmic problems
faced in the field and how modern CS research can contribute to their
solution.

As noted above, in this article we choose to focus on cheminformatics
in the context of drug discovery and development. Given that this is a
slow and expensive process, the concept of ``risk minimization'' plays
an important role in this field. Thus, we structure the article around
this idea -- how to minimize the chances of a small
molecule failing, due to poor physical, chemical or biological
properties, in the various research and development stages as a drug
candidate, and how to successfully employ cheminformatics methods in
these efforts. As an example, a molecule must be soluble and show a
certain degree of bioavailability to be considered as drug
candidate. In cases where these properties are poor, a cheminformatics
approach can suggest replacement of certain functional groups, that
maintain potency but improve the solubility and bioavailability.
Table~\ref{table:properties} (in the Appendix) summarizes a number of
properties a drug candidate must satisfy to be considered
therapeutically useful (such as being appropriately soluble,
exhibiting minimal toxicity and being selective for the intended
target) and sketches the role of cheminformatics at each stage of drug
development.
%
\subsection{Representing and Searching Structures}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical
structures, their properties, and relationships to, \textit{e.g.},
biological targets.  Organizing and maintaining such databases, as
well as searching and clustering similar structures together, are
essential to enable many scientific applications. Yet, each one of
these areas poses its unique computer science challenges.

@@

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine-readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. Considering protonation states 
(how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge  
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

In the quest for the discovery of novel therapeutic agents, it is
essential to study molecules as they are situated in a biological
context, as metabolites, cofactors for enzymatic reactions, hormones
and signalling molecules. Thus we need to adequately represent and
organise chemical data across biological databases such as those
containing pathways, protein information or biological
samples. Integration and processing of data from such disparate
domains underlies ``systems'' level biological research.  This leads
to additional challenges that standard chemical structural
representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class
of compounds behave in a certain context in a biological system rather
than a single molecule with a fully-specified structure -- and for
similar reasons, chemical concepts such as \textit{groups} (parts of
molecules); secondly, the need to represent \textit{non-structural}
groupings of compounds of interest, since compounds may bind to the
same or similar targets or they may have similar biological functions
such as acting as an anti-neoplastic -- preventing the development of
tumors. With the many different biological databases needing to refer
to these different chemical entities in an organised and standardized
fashion to enable cross-database integration for a whole-systems
perspective, \textit{ontologies} and other semantic technologies are
being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within biological
systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases.
Ontologies are based on logical languages of varying levels of
expressivity, accompanied by sophisticated reasoning algorithms. They
can be used to construct various sorts of semantic similarity
(e.g. similarity in function, similarity in application) which
complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical structure
information with the wider logic-based ontology information, which
will allow combined automated reasoning and querying over the
different domains encoded in a growing number of interlinked
ontologies. % \cite{hastingsowled2010}.
Enabling automated inferencing in such heterogenous linked systems
brings us closer to achieving a systems level understanding of small
molecule activity \cite{Oprea:2007fk}.

@@

Another application of structure enumeration is in the area of Markush
structures and searches \cite{Barnard:1991vn}, where the user
specifies a pattern such as ``\emph{\ldots an aromatic ring with two
  alkyl groups attached to it}.'' Such a pattern is very general: an
alkyl group can be methyl, ethyl or any chain of $n$ carbons and there
are three possible positions for them on the ring. Even in this simple
example there are $3n^2$ possible structures. For more complex
Markushes, there can be billions of possible structures. Clearly,
explicit enumeration is not feasible. Thus one challenge is to devise
methods to search (based on structural or property similarity) through
the implicit enumerated space. A number of commerical vendors such as
Digitial Chemistry and ChemAxon provide toolkits to handle these
problems.

Structure enumeration plays a fundamental role in \emph{molecular
  design} -- the design of compounds (drugs, for instance) that
optimize some physical, chemical, or biological property or activity
\cite{Schneider:2005uq}. A key challenge in this area is to combine
enumeration algorithms with efficient property prediction and is
closely related to methods in predictive modeling of chemical
properties (Section \ref{sec:activity-mining}). This problem is also
termed \emph{inverse QSAR}, where one must devise a set of features
that describe a molecule with a specified property, and subsequently
reconstruct the molecule from those features. As a side note, the
reconstruction of molecules from feature sets is can be considered a
cryptography problem. It is relevant to the pharmaceutical industry,
as it can be necessary to share data on molecules without sharing
explicit structures. Feature sets that allow easy reconstruction of
molecular structures are thus undesirable. A number of workers have
investigated the problem of identify such one-way molecular
descriptors as well as methods to reconstruct molecules from
descriptors \cite{Masek:2008kx} and new methods that generate such
one-way molecular features but still correlate with physicochemical
properties would be valuable.

@@

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
 Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking (which predicts how molecules fit together),
pharmacophore modeling (a 3D approach to capturing protein-ligand
interactions) and proteochemometric methods have been designed to take
into account both ligand and receptor structures. The last method is
an extension of statistical QSAR methods to simultaneously
model the receptor and ligand, as first reported by Lapinsh \textit{et
 al.}~\cite{lapinsh2001}.

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. Here
multiple approaches exist. We can transform molecular graphs into a
numerical vector representation of chemical structures
\cite{todeschini2000}, ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vectorial
representations, we directly operate on the graph representations.

Both methods have advantages and disadvantages. The vectorial approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors: the feature selection problem 
well discussed in the data mining literature. A kernel approach
does not require feature selection, but one faces the problem of
evaluating a data set in a pairwise fashion, and one must
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a
well-known mathematical property in machine learning that makes a set
of observations easier to make predictions about), and this is not
always possible with traditional cheminformatics based kernels (such
as a multiple common substructure based kernel). These challenges can
make kernel based methods prohibitive on very large datasets.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic (for example)? The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and fast
predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and Partial Least Squares) though
some algorithmic approaches such as random forests can provide insight.

@@

At this point, ``electronic lab notebooks have replaced the
traditional paper lab book across the pharmaceutical industry,'' but
are not yet common in academia, because of their high
cost~\cite{WavingGoodbye}.  It is worth mentioning that chemistry is
traditionally a field where many people have been accustomed to huge
bookshelves of reactions and compounds. Chemists changed to online
ELNs not only to increase efficiency, but also because it allowed easy
collaboration with trusted colleagues. Third party paper catalogs and
academic publications proved to be inefficient, and even more
critically, they did not take compounds of trusted colleagues into
account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required delivering on the promise to increase
interconnectivity, and to improve collaboration opportunities.  The
leading players in this area include CambridgeSoft's E-Notebook and
Accelrys' Symyx Notebook, but there are at least 35 different
companies producing ELNs as of February 2011~\cite{ELNreview}, for use
in chemistry, biology, quality control, and other research domains.  Another
knowledge management tool specifically for chemists is the web-based
service Reaxys, founded in 2009, which aims to provide ``a
single, fully integrated chemical workflow solution.''

However, the future still leaves many challenges ahead. Using
external/public databases with chemical and bioactivity data remains a
challenge due to differences in identifiers, synchronization, curation
\& error correcting mechanisms and efficient substructure and
similarity search within complex data types (compounds). A
collaboration with external parties, e.g. contract research, poses
other problems including compound duplication and efficient data
synchronization. If external partners are small they often do not even
have sufficient IT resources themselves and thus rely on external
services. Cloud services could help not only to provide a service
infrastructure for all parties involved, but also to provide the
required private/public interface. Furthermore, there are still many
data sources that are not being indexed properly: chemistry patents
are often cryptic and chemical image and text mining remains a
challenge (albeit, currently being addressed in academia and industry
\cite{Jessop:2011fk,Sayle:2009uq}). The closed CAS (Chemical Abstract
Service) is a highly trusted source, and public chemistry and
bioactivity space has to improve quality and interconnectivity to
compare with it. Efforts like ChEMBL and PubChem-BioAssay are on the
right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring
constant growth, usage, and collaboration scenarios between private
and public parties.

@@

The extent of Open Source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly due to the
existence of a lucrative pharmaceutical market for proprietary
software. However, in recent years there has been a rise in the amount
of Open Source cheminformatics software. One contribution to this has
been the Blue Obelisk group~\cite{BlueObelisk2011}
(\url{http://blueobelisk.org}) of chemists, who have worked together
to create interoperable Open Source chemistry software (among other
activities). Now there are a number of Open Source chemical toolkits
available including the Chemistry Development Kit
(CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and
Indigo. These toolkits are written in Java or C++ (though toolkits
written in the latter invariably have bindings to a variety of
languages). Such toolkits are used, for example, to read/write
chemical file formats, manipulate chemical structures, measure
similarity of molecules, search for substructures, and generate 2D
diagrams. The underlying algorithms used include graph algorithms
(e.g. maximal common substructure, canonical labelling of graphs),
geometrical methods (e.g. Kabsch alignment) and vector manipulation
(e.g. converting coordinates in various systems, 3D structure
generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse~\cite{Bioclipse2} uses the CDK.

@@

There is an admittedly steep learning curve, as significant domain
knowledge is required to attack many problems in cheminformatics due
to the underlying chemistry of the systems being studied. Indeed, many
issues faced by chemists do not admit as clean an abstraction as
``algorithms on a string'', the way many bioinformatics algorithms can
be abstracted into a computer science framework.  As a result, while
one can certainly make significant contributions to cheminformatics by
just considering structures as graphs, one is limited to somewhat
abstract problems. Keeping in mind that cheminformatics is
fundamentally a practical field that serves to help and advance
experimental chemistry, the key challenges require an understanding of
the underlying chemical systems.

>>>>

\title{Cheminformatics, the Computer Science of Chemical Discovery, Turning Open-Source}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Advancing Translational Science}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

@@

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: the
  interdisciplinary collaboration between computer scientists and
  molecular biologists that led as one of its major accomplishments to the
  sequencing of the human genome. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. While researchers in both academia and industry have made important contributions to this field since the 1960s (if not before), new and exciting research opportunities now exist because of an ``opening'' of data and software that began around the year 2000.  Since that time -- as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, as a result of chemists volunteering their time
  for an Open Science ``movement'' -- researchers have gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that might have been prohibitive to tackle a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Novel technologies in the life sciences produce information at an
ever increasing rate -- public data stores such as that of the
European Bioinformatics Institute (EBI) contain in the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available software implementing many of
these algorithms has only been accessible in the last ten to fifteen
years \cite{faulon2010}.

@@

Given that the art and science of drug discovery is based around small
molecules, techniques and algorithms to handle, analyse and interpret
chemical structure information are a core component of the methods of drug
discovery. Public availability of chemical information and
tools to handle it are important as drug discovery research is slow,
expensive and prone to failure. The more information available to each
researcher, facilitating the development of novel treatments, the
better the chances of avoiding the many causes of attrition. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
\emph{annotations} (comments on the structure that are database
searchable), such as synonyms (different names for the same molecule),
known targets (drugs with which the molecule is known to interact),
mode of action (how the molecule interacts with its targets),
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

@@

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The representations range
from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
much implicit information (such as unlabeled nodes represent carbons
with valencies satisfied by hydrogens) to a fully labeled graph,
where all information is explicit
(Fig.~\ref{figure:chemical-structures}B). The former is what is
usually exchanged informally between chemists, but to actually mine
chemical structures computationally, something like the latter is
required.  For example, graph mining algorithms can be applied to
labeled graph representations, to discover patterns in large-scale
datasets \cite{horst2009}; this information can also be determined by simulations such as molecular dynamics.

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (top left), a labeled graph (top right), a 3D
  conformation (bottom left) and a surface representation (bottom right). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes, with the choice of representation guided by its
  intended use.}
\label{figure:chemical-structures}
\end{figure}

@@

This main article will highlight current topics in
cheminformatics, presenting the computational and algorithmic problems
faced in the field and how modern CS research can contribute to their
solution.  We have also provided two appendices: a brief history of cheminformatics, and a detailed list of open problems.

We structure the main article around ``risk minimization'' of drug discovery (a slow and expensive process)---how to minimize the chances of a small
molecule failing, due to poor physical, chemical or biological
properties, in the various research and development stages as a drug
candidate, and how to successfully employ cheminformatics methods in
these efforts. As an example, a molecule must be soluble and show a
certain degree of bioavailability to be considered as a drug
candidate. In cases where these properties are poor, a cheminformatics
approach can suggest replacement of certain functional groups, that
maintain potency but improve the solubility and bioavailability.
Table~\ref{table:properties} (in the Appendix) summarizes a number of
properties a drug candidate must satisfy to be considered
therapeutically useful (such as being appropriately soluble,
exhibiting minimal toxicity and being selective for the intended
target) and sketches the role of cheminformatics at each stage of drug
development.
%
\subsection{Representing and Searching Structures}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical
structures, their properties, and relationships to, \textit{e.g.},
biological targets.  Organizing and maintaining such databases, as
well as searching and clustering similar structures together, are
essential to enable many scientific applications. Yet, each one of
these areas poses its unique computer science challenges.

@@

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine-readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. Considering protonation states
(how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format as defined by Weininger 1988\cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

In the quest for the discovery of novel therapeutic agents, it is
essential to study molecules as they are situated in a biological
context, as metabolites, cofactors for enzymatic reactions, hormones
and signalling molecules. Thus we need to adequately represent and
organise chemical data across biological databases such as those
containing pathways, protein information or biological
samples. Integration and processing of data from such disparate
domains underlies ``systems'' level biological research.  This leads
to additional challenges that standard chemical structural
representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class
of compounds behave in a certain context in a biological system rather
than a single molecule with a fully-specified structure -- and for
similar reasons, chemical concepts such as \textit{groups} (parts of
molecules); secondly, the need to represent \textit{non-structural}
groupings of compounds of interest, since compounds may bind to the
same or similar targets or they may have similar biological functions
such as acting as an anti-neoplastic -- preventing the development of
tumors. With the many different biological databases needing to refer
to these different chemical entities in an organized and standardized
fashion to enable cross-database integration for a whole-systems
perspective, \textit{ontologies} and other semantic technologies are
being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within biological
systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases.
Ontologies are based on logical languages of varying levels of
expressivity, accompanied by sophisticated reasoning algorithms. They
can be used to construct various sorts of semantic similarity
(e.g. similarity in function, similarity in application) which
complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical structure
information with the wider logic-based ontology information, which
will allow combined automated reasoning and querying over the
different domains encoded in a growing number of interlinked
ontologies. % \cite{hastingsowled2010}.
Enabling automated inferencing in such heterogenous linked systems
brings us closer to achieving a systems level understanding of small
molecule activity \cite{Oprea:2007fk}.

@@

Another application of structure enumeration is in the area of Markush
structures and searches \cite{Barnard:1991vn}, where the user
specifies a pattern such as ``\emph{\ldots an aromatic ring with two
  alkyl groups attached to it}.'' Such a pattern is very general: an
alkyl group can be methyl, ethyl or any chain of $n$ carbons and there
are three possible positions for them on the ring. Even in this simple
example there are $3n^2$ possible structures. For more complex
Markushes, there can be billions of possible structures. Clearly,
explicit enumeration is not feasible. Thus one challenge is to devise
methods to search (based on structural or property similarity) through
the implicit enumerated space. A number of commercial vendors such as
Digitial Chemistry and ChemAxon provide toolkits to handle these
problems.

Structure enumeration plays a fundamental role in \emph{molecular
  design} -- the design of compounds (drugs, for instance) that
optimize some physical, chemical, or biological property or activity
\cite{Schneider:2005uq}. A key challenge in this area is to combine
enumeration algorithms with efficient property prediction and is
closely related to methods in predictive modeling of chemical
properties (Section \ref{sec:activity-mining}). This problem is also
termed \emph{inverse QSAR}, where one must devise a set of features
that describe a molecule with a specified property, and subsequently
reconstruct the molecule from those features. As a side note, the
reconstruction of molecules from feature sets is can be considered a
cryptography problem. It is relevant to the pharmaceutical industry,
as it can be necessary to share data on molecules without sharing
explicit structures. Feature sets that allow easy reconstruction of
molecular structures are thus undesirable. A number of workers have
investigated the problem of identifying such one-way molecular
descriptors as well as methods to reconstruct molecules from
descriptors \cite{Masek:2008kx} and new methods that generate such
one-way molecular features but still correlate with physicochemical
properties would be valuable.

@@

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
 Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking (which predicts how molecules fit together),
structure-based pharmacophore modeling (a 3D approach to capturing protein-ligand
interactions) and proteochemometric methods have been designed to take
into account both ligand and receptor structures. The last method is
an extension of statistical QSAR methods to simultaneously
model the receptor and ligand, as first reported by Lapinsh \textit{et
 al.}~\cite{lapinsh2001}.

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. Here
multiple approaches exist. We can transform molecular graphs into a
numerical vector representation of chemical structures
\cite{todeschini2000}, ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vector
representations, we directly operate on the graph representations.

Both methods have advantages and disadvantages. The vector approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors: the feature selection problem is
well discussed in the data mining literature. A kernel approach
does not require feature selection, but one faces the problem of
evaluating a data set in a pairwise fashion, and one must
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a
well-known mathematical property in machine learning that makes a set
of observations easier to make predictions about), and this is not
always possible with traditional cheminformatics based kernels (such
as a multiple common substructure based kernel). These challenges can
make kernel based methods prohibitive on very large datasets.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic (for example)? The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and fast
predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (e.g., neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and Partial Least Squares) though
some algorithmic approaches such as random forests can provide insight.

@@

At this point, ``electronic lab notebooks have replaced the
traditional paper lab book across the pharmaceutical industry,'' but
are not yet common in academia, because of their high
cost~\cite{WavingGoodbye}.  It is worth mentioning that chemistry is
traditionally a field where many people have been accustomed to huge
bookshelves of reactions and compounds. Chemists changed to online
ELNs not only to increase efficiency, but also because it allowed easy
collaboration with trusted colleagues. Third party paper catalogs and
academic publications proved to be inefficient, and even more
critically, they did not take compounds of trusted colleagues into
account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required delivering on the promise to increase
interconnectivity, and to improve collaboration opportunities.  These opportunities were made legally possible in the USA because of FDA regulation Title 21 CFR Part 11, which permits the use of electronic signatures in chemical documents.  The
leading players in this area include CambridgeSoft's E-Notebook and
Accelrys' Symyx Notebook, but there are at least 35 different
companies producing ELNs as of February 2011~\cite{ELNreview}, for use
in chemistry, biology, quality control, and other research domains.  Another
knowledge management tool specifically for chemists is the web-based
service Reaxys, founded in 2009, which aims to provide ``a
single, fully integrated chemical workflow solution.''

However, the future still leaves many challenges ahead. Using
external/public databases with chemical and bioactivity data remains a
challenge due to differences in identifiers, synchronization, curation
\& error correcting mechanisms and efficient substructure and
similarity search within complex data types (compounds). A
collaboration with external parties, e.g. contract research, poses
other problems including compound duplication and efficient data
synchronization. If external partners are small they often do not even
have sufficient IT resources themselves and thus rely on external
services. Cloud services could help not only to provide a service
infrastructure for all parties involved, but also to provide the
required private/public interface. Furthermore, there are still many
data sources that are not being indexed properly: chemistry patents
are often cryptic and chemical image and text mining remains a
challenge (albeit, currently being addressed in academia and industry
\cite{Jessop:2011fk,Sayle:2009uq}). The closed CAS (Chemical Abstract
Service) is a highly trusted source, and public chemistry and
bioactivity space has to improve quality and interconnectivity to
compare with it. SciFinder is another chemical abstract service, with a version, SciFinder Scholar, marketed to universities.  Open-source efforts like ChEMBL and PubChem-BioAssay are on the
right track, though, unlike some commercial tools, these efforts do not yet abstract reactions. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring
constant growth, usage, and collaboration scenarios between private
and public parties.

@@

The extent of Open Source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly due to the
existence of a lucrative pharmaceutical market for proprietary
software. However, in recent years there has been a rise in the amount
of Open Source cheminformatics software. One contribution to this has
been the Blue Obelisk group~\cite{BlueObelisk2011}
(\url{http://blueobelisk.org}) of chemists, who have worked together
to create interoperable Open Source chemistry software (among other
activities). Now there are a number of Open Source chemical toolkits
available including the Chemistry Development Kit
(CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and
Indigo. These toolkits are written in Java or C++ (though toolkits
written in the latter invariably have bindings to a variety of
languages). Such toolkits are used, for example, to read/write
chemical files in various formats, manipulate chemical structures, measure
similarity of molecules, search for substructures, and generate 2D
depictions. The underlying algorithms used include graph algorithms
(e.g. maximal common substructure, canonical labelling of graphs),
geometrical methods (e.g. Kabsch alignment) and vector manipulation
(e.g. converting coordinates in various systems, 3D structure
generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse~\cite{Bioclipse2} uses the CDK.

@@

There is an admittedly steep learning curve, as significant domain
knowledge is required to attack many problems in cheminformatics due
to the underlying chemistry of the systems being studied. Indeed, many
issues faced by chemists do not admit as clean an abstraction as
``algorithms on a string,'' the way many bioinformatics algorithms can
be abstracted into a computer science framework.  Hence, while
one can certainly make significant contributions to cheminformatics by
just considering structures as graphs, one is limited to somewhat
abstract problems. Keeping in mind that cheminformatics is
fundamentally a practical field that serves to help and advance
experimental chemistry, the key challenges require an understanding of
the underlying chemical systems.
